Discovery of the first quadruple gravitationally lensed quasar candidate with Pan-STARRS

### 2017-05-14 paper4.tex
finished abstract, section 4 and plots

\shorttitle{A quadruple gravitationsal lens in PS1}
\shortauthors{C. Berghea et al.}

@@

\title{ Discovery of the first quadruple gravitationsally lensed quasar candidate with Pan-STARRS }

@@

\altaffiltext{1}{U.S. Naval Observatory (USNO), 3450 Massachusetts Avenue NW, Washington, DC 20392, USA}
\altaffiltext{2}{Department of Physics, University of California, Davis, CA 95616, USA}
\altaffiltext{3}{Rutgers, the State University of New Jersey, 136 Frelinghuysen Road, Piscataway, NJ 08854, USA}

@@

We report the serendipitous discovery of the first quadruple gravitational lens candidate in the Pan-STARRS images. The lensing galaxy is detected by fitting the lens in i, z and y bands using the Hostlens software. There are small differences in the colors of the four quasars images but they are generally consistent quasar at about z $\sim$ 2.6 based on photometric redshift estimates. The lensing galaxy is very faint but we used galaxy templates to estimate z $\sim$ 0.6. We model the lens with an isothermal ellipsoid (SIE) with shear. While both models provide relatively good fits to the data, there are important discrepancies in the fluxes, especially betwen the B and C components which the models predict should show similar fluxes. These can be explained by differential extinction, microlensing by stars and millilensing due to the dark matter substructure. 
\end{abstract}

@@

We assume a cosmological model with $\Omega_M = 0.274$, $\Omega_L = 0.726$, and h = 0.71. 
In Section 2 we present the results from relative astrometry, photometry and photometric redshift estimates. In Section 3 we model the lens with the gravlens software. Discussion and conclusion are finally presented in Section 4.

@@

\section{CONCLUSIONS}

The evidence is overwhelming for this to be a lens (configuration reproduced by lensing, similar color for each image, detection of lensing galaxy), but we stop short of claiming that it is a confirmed lens until we get spectroscopy.

@@

\begin{deluxetable}{l|ccccc}
\tablecolumns{5}  
\tablewidth{0pt} 
\tabletypesize{\tiny}
\setlength{\tabcolsep}{0.05in}
\tablenum{1} %\\               
\tablecaption{Lens measurements and modeling results}   
\tablehead{
\colhead{Property} & \colhead{A} & \colhead{B} & \colhead{C} & \colhead{D} & \colhead{G}\\
}
\startdata

@@

g & 15.60$\pm$0.01 &  15.72$\pm$0.01 &  16.45$\pm$0.02 & 18.09$\pm$0.01 & $\cdots$ \\
r & 15.40$\pm$0.01 &  15.55$\pm$0.01 &  16.21$\pm$0.01 & 17.74$\pm$0.01 & $\cdots$ \\
i & 15.36$\pm$0.01 &  15.57$\pm$0.02 &  16.15$\pm$0.02 & 17.74$\pm$0.02 & 19.50$\pm$0.20 \\
z & 15.23$\pm$0.03 &  15.50$\pm$0.05 &  16.02$\pm$0.01 & 17.68$\pm$0.03 & 18.95$\pm$0.13 \\
y & 14.92$\pm$0.01 &  15.23$\pm$0.02 & 15.76$\pm$0.02 & 17.36$\pm$0.02 & 19.20$\pm$0.24 \\
$\Delta$i & 0.0 & 0.21 & 0.79 & 2.38 & 4.14 \\
$\Delta$x & 0.000$\pm$0.004 & -1.185$\pm$0.004 & 1.271$\pm$0.005 & 0.410$\pm$0.004 & 0.240$\pm$0.050 \\
$\Delta$y & 0.000$\pm$0.004 & -0.441$\pm$0.004 & -0.074$\pm$0.004 & -3.310$\pm$ 0.004 & -2.310$\pm$0.025 \\
z & 0.818$^{+0.007}_{-0.009}$ & 0.732$^{+0.007}_{-0.006}$ & 0.912$^{+0.008}_{-0.009}$ & 0.914$^{+0.007}_{-0.009}$  \\

@@

\tablecomments{
The first four columns are the four quasar images and the fifth column is the lens.
The first five rows are magnitude measurements derived using Hostlens and corrected using NED extinction calculator.
The following three rows are relative i magnitudes and relative positions in arcseconds averaged between the i and y bands. 
The next row shows the photometric redshift estimates following \citet{wu10}.
Following rows shows the model prediction for position and relative magnitudes for two gravitational models.
}
\end{deluxetable}

\begin{figure}
\epsscale{1.1}
\plottwo{lens3.eps}{lens4.eps}
\caption{
PS1 images of the lens candidate. Left: is a color coded close image R-y, G-i, B-g. Right: y image of the surrounding area, image size is 2 arcmin. The bright star south-east of the lens is saturated in the other bands.
}
\label{lens}
\end{figure}

\begin{figure}
\epsscale{1.1}
\plottwo{redshiftseparate.eps}{redshifttogether.eps}
\caption{
Photometric redshift estimates. 
Left: individually for each image; right: for all images simultanesously}
\label{redshift}
\end{figure}

\begin{figure}
\epsscale{1.1}
\plotone{hostlens.eps}
\caption{
Hostlens i band fits. From left to right: the original PS1 image, after the quasar images are removed and finally after the lens image is also removed}
\label{hostlensi}
\end{figure}

\begin{figure}
\epsscale{0.6}
\plotone{critcrv1.eps}
\caption{
Gravitational lens model. The red and blue lines show the critical curve and caustics. The green dot shows the quasar position and the black dots the lensed image position}
\label{model}
\end{figure}

\begin{figure}
\epsscale{1.0}
\plotone{modelphotometry.eps}
\caption{
Comparison between the observed quasar image magnitudes differences (vertical colored lines) and the model predicitons (blue) for each PS1 band. The model predicitons were drrawn from a MCMC to explore the range of models. The discrepancied between the model and observed are clearly visible}
\label{modelphot}
\end{figure}

>>>>

\shorttitle{A quadruply lensed quasar in PS1}
\shortauthors{C. Berghea et al.}

@@

\title{ Discovery of the first quadruple gravitationally lensed quasar candidate with Pan-STARRS }

@@

\altaffiltext{1}{U.S. Naval Observatory (USNO), 3450 Massachusetts Avenue NW, Washington, DC 20392, USA}
\altaffiltext{2}{Department of Physics, University of California, Davis, 1 Shields Avenue, CA 95616, USA}
\altaffiltext{3}{Rutgers, the State University of New Jersey, 136 Frelinghuysen Road, Piscataway, NJ 08854, USA}

@@

We report the serendipitous discovery of the first gravitationally lensed quasar candidate from Pan-STARRS. The $grizy$ images reveal four point-like images with magnitudes between $14.9$ mag and $18.1$ mag. The colors of the point sources are similar, and they are more consistent with quasars than stars or galaxies. The lensing galaxy is detected in the $izy$ bands, with an inferred photometric redshift of $\sim 0.6$, lower than that of the point sources. We successfully model the system with a singular isothermal ellipsoid with shear, using the relative positions of the five objects as constraints. While the brightness ranking of the point sources is consistent with that of the model, we find discrepancies between the model-predicted and observed fluxes, likely due to microlensing by stars and millilensing due to dark matter substructure. In order to fully confirm the gravitational lens nature of this system, and add it to the small but growing number of the powerful probes of cosmology and astrophysics represented by quadruply lensed quasars, we further require spectroscopy and high-resolution imaging.

\end{abstract}

@@

We assume a cosmological model with $\Omega_M = 0.274$, $\Omega_L = 0.726$, and h = 0.71. All magnitudes are in the AB system. The structure of the paper is as follows:
In Section 2 we infer relative astrometry, photometry and photometric redshifts. In Section 3 we fit a mass model to the system. Conclusions and future works are presented in Section 4.

@@

\section{CONCLUSIONS AND FUTURE WORKS}

We have presented the first quadruply imaged gravitationally lensed quasar candidate in the Pan-STARRS1 Survey, discovered via visual inspection of the multi-band images. We find that the evidence supporting the gravitational lens nature of this system is overwhelming, and consists of the following:

\begin{itemize}
\item The presence of four point-like images within $\lesssim 3.8\arcsec$ of eachother, with similar colors, completely consistent with quasars templates, but less consistent with stars and galaxies.
\item The detection of a red galaxy between the point source, such that the relative positions of all five objects are fully consistent with a well-known ``cusp'' configuration. We successfully reproduce this configuration using a SIE + shear model for the lensing galaxy. Further more, the brightness ranking of the point-like images is consistent with the ones predicted by the model.
 \item The photometric redshift analysis shows that the galaxy is of elliptical template, typical for lensing galaxies, and of redshift smaller than the point sources. For the inferred redshifts, the velocity dispersion and mass of the lensing galaxy are consistent with those of typical galaxies.
\end{itemize}

Nevertheless, we stop short of claiming that this is a confirmed lens until we can obtain spectroscopy of this system. Spectroscopy of the four point-like images would unequivocally demonstrate whether they are multiple images of the same background quasar, as well as allow to infer its redshift and physical properties. We also note that the system is accessible to ground based high-resolution observations at most adaptive optics-capable facilities in the Northern hemisphere, due to the proximity to a $R\sim12$ mag star $\sim18\arcsec$ away.

@@

\begin{deluxetable}{l|ccccc}
\tablecolumns{5}  
\tablewidth{0pt} 
\tabletypesize{\tiny}
\setlength{\tabcolsep}{0.05in}
\tablenum{1} %\\               
\tablecaption{Relative astrometry and photometry}   
\tablehead{
\colhead{Property} & \colhead{A} & \colhead{B} & \colhead{C} & \colhead{D} & \colhead{G}\\
}
\startdata

@@

g & 15.60$\pm$0.01 &  15.72$\pm$0.01 &  16.45$\pm$0.02 & 18.09$\pm$0.01 & $\cdots$ \\
r & 15.40$\pm$0.01 &  15.55$\pm$0.01 &  16.21$\pm$0.01 & 17.74$\pm$0.01 & $\cdots$ \\
i & 15.36$\pm$0.01 &  15.57$\pm$0.02 &  16.15$\pm$0.02 & 17.74$\pm$0.02 & 19.50$\pm$0.20 \\
z & 15.23$\pm$0.03 &  15.50$\pm$0.05 &  16.02$\pm$0.01 & 17.68$\pm$0.03 & 18.95$\pm$0.13 \\
y & 14.92$\pm$0.01 &  15.23$\pm$0.02 & 15.76$\pm$0.02 & 17.36$\pm$0.02 & 19.20$\pm$0.24 \\
%$\Delta$i & 0.0 & 0.21 & 0.79 & 2.38 & 4.14 \\
$\Delta$x & 0.000$\pm$0.004 & -1.185$\pm$0.004 & 1.271$\pm$0.005 & 0.410$\pm$0.004 & 0.240$\pm$0.050 \\
$\Delta$y & 0.000$\pm$0.004 & -0.441$\pm$0.004 & -0.074$\pm$0.004 & -3.310$\pm$ 0.004 & -2.310$\pm$0.025 \\
%photoz & 0.818$^{+0.007}_{-0.009}$ & 0.732$^{+0.007}_{-0.006}$ & 0.912$^{+0.008}_{-0.009}$ & 0.914$^{+0.007}_{-0.009}$  \\

@@

\tablecomments{
The first four columns are the four quasar images and the fifth column is the lens.
The first five rows are magnitude measurements derived using Hostlens, corrected for extinction.
%The following three rows are relative i magnitudes and relative positions in arcseconds averaged between the i and y bands. 
%The next row shows the photometric redshift estimates following \citet{wu10}.
The last raw shows the model prediction for the relative magnitudes relative to image A, from the best-fit mass model.
}
\end{deluxetable}

\begin{figure}
\epsscale{1.1}
\plottwo{lens3.eps}{lens4.eps}
\caption{
PS1 images of the lens candidate. North is up and east is to the left. Left: close-up color image using the $g$ (blue), $i$ (green) and $y$ (red) filters. Right: $y$-band image of a 2 $\arcmin$ region around the system. The bright star south-east of the lens is saturated in the other bands.}
\label{lens}
\end{figure}

\begin{figure}
\epsscale{1.1}
\plottwo{redshiftseparate.eps}{redshifttogether.eps}
\caption{
Photometric redshift estimates using the method in \citet{wu10}. 
Left: individually for each image; right: for all images simultanesously.}
\label{redshift}
\end{figure}

\begin{figure}
\epsscale{1}
\plotone{hostlens.eps}
\caption{
Original images ($10\arcsec\times10\arcsec$, first row) and residuals after subtracting the best-fit model with Hostlens. From left to right: $g$, $r$, $i$, $z$, $y$-band. In the second raw, the lensing galaxy is not including in the modeling, whereas in the third raw it is. The images are in linear scale, and cover the full dynamic range.}
\label{hostlensi}
\end{figure}

\begin{figure}
\epsscale{0.6}
\plotone{critcrv1.eps}
\caption{
Gravitational lens modelling. The blue and red lines show the critical curves and caustics, respectively. The green dot shows the quasar position in the source plane, and the black dots show the lens and the observed image positions.}
\label{model}
\end{figure}

\begin{figure}
\epsscale{1.0}
\plotone{model-photometry.pdf}
\caption{
Comparison between the observed quasar image magnitudes relative to image A (vertical colored lines) and the model predictions (blue histogram) for each PS1 band. The model predictions were drawn from a MCMC exploration of the range of models}
\label{modelphot}
\end{figure}

\begin{figure}
\epsscale{1.1}
\plotone{ilens_out_file_mcmc.png}
\caption{
Degeneracies between the parameters used to model the observed system in $i$-band with Hostlens. The three contour lines mark the 1-, 2- and 3-$\sigma$ limits. The vertical lines in the histograms mark the 16th, 50th and 84th percentiles. FWHMs are given in arcseconds, positions in pixels, and fluxes in counts.}
\label{mcmc_hostlens}
\end{figure}
